---
title: "Masters Homework 2"
author: "Nouf Alharbi"
date: "7/21/2025"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, echo = TRUE}

nyc <- read.csv("https://raw.githubusercontent.com/fredlapolla/RVilcekMaster23/master/NYC_HANES_DIAB.csv")

  
```

>- Find summary information about the total blood mercury levels in our NYC HANES data (hint: names(nyc) will call the names of each column). What is the max value, what is the minimum value.  

```{r}
names(nyc)

summary(nyc$MERCURYU)
max(nyc$MERCURYU, na.rm = TRUE)
min(nyc$MERCURYU, na.rm = TRUE)

```

>- Bonus/Challenge: What is the Standard Deviation and skewness of the blood mercury?

```{r}
sd_mercury <- sd(nyc$MERCURYU, na.rm = TRUE)
cat("The standard deviation of blood mercury is", sd_mercury)

install.packages("moments")
library(moments)

sk_mercury <- skewness(nyc$MERCURYU, na.rm = TRUE)
cat("The skewness of blood mercury is", sk_mercury)


```



>- Install a package and load, such as "ggplot2" or "dplyr"

```{r}
install.packages("ggplot2")
library(ggplot2)

```



>- What two commands are needed to install and use a package in R?

```{r}
# install.packages("package_name")
# library(package_name)
```




>- When you read in a .csv file into R, what class is the object that you read in (like nyc)?
```{r}
class(nyc)
```


>- What command can you use to determine what type of data is in a column?

```{r}
#class(data.frame$column)
```


>- Provide a command to find the type of data in the DX_DBTS column of nyc. 

```{r}
class(nyc$DX_DBTS)
```


>- Provide the same command the type of data but use the other way of indexing from what you did above. 

```{r}

class(nyc[["DX_DBTS"]])

```



>- What would happen if we had chosen the first 4 columns of NYC HANES and made a matrix? What would the class of the cells in this matrix be? Why?

```{r}
nyc_matrix <- as.matrix(nyc[,1:4])
class(nyc_matrix)
class(nyc_matrix[1,1])
class(nyc_matrix[2,])
# The class of the cells in the matrix would be "character", because matrices cannot hold different types of data so it converted them all into character strings because that is the lowest data type in the hierarchy.
```


>- If we make a vector,vec<- c(3,9,4,7,8,9), and multiply it by WannaMatrix, what is the value in the third row, third column? 

```{r}
WannaMatrix <- nyc[5:10, 6:9]
WannaMatrix <- as.matrix(WannaMatrix,nrow = 6, ncol = 4)

vec <- c(3,9,4,7,8,9)

new_mat <- WannaMatrix * vec
new_mat[3,3]

```



>- What happens if instead we made a vector vec2 <- c(3,9,4,7,8) and multiply it by WannaMatrix, what is the third row, third column value? What is going on? 

```{r}
vec2 <- c(3,9,4,7,8)

length(vec2)
dim(WannaMatrix)

new_mat2 <- WannaMatrix * vec2

#third row, third column value
new_mat2[3,3]

#checking dimension of new matrix
dim(new_mat2)

#since vec2 has 5 elements, and WannaMatrix has 5 rows, this warning messages pops up: 
#  Warning in WannaMatrix * vec2 :
# longer object length is not a multiple of shorter object length
# vec2 will use the first element as the 6th, ie vec2 will become (3,9,4,7,8,3) to match the dimensions of the matrix
```


>- Look up the arguments for read.csv

```{r}

?read.csv

```


>- Use apply() to find the median value of CADMIUM, LEAD CHOLESTEROLTOTAL, GLUCOSESI, CREATININESI, and CREATININE

```{r}

apply(nyc[, c("CADMIUM", "LEAD", "CHOLESTEROLTOTAL", "GLUCOSESI", "CREATININESI", "CREATININE")], 2, median, na.rm = TRUE)

```


>- Do the same using sapply()

```{r}

sapply(nyc[, c("CADMIUM", "LEAD", "CHOLESTEROLTOTAL", "GLUCOSESI", "CREATININESI", "CREATININE")], median, na.rm = TRUE)


```

>- Use a for loop to find the median value of columns 16:19

```{r}

for(i in 16:19) {
  print(median(nyc[[i]], na.rm = TRUE))
}

```


>- Make a for loop to find the standard deviation (hint sd() ) of CHOLESTEROLTOTAL, GLUCOSESI, CREATININESI, AND CREATININE

```{r}
cols <- c("CHOLESTEROLTOTAL", "GLUCOSESI", "CREATININESI", "CREATININE")

for (col in cols) {
  sd_cols <- sd(nyc[[col]], na.rm = TRUE)
   print(sd_cols)
  cat("The standard deviation of", col, "is", sd_cols, "\n")
}
```


>- Bonus/Challenge Create a vector in which to put the standard deviations 

```{r}

sd_vec <- numeric(length(cols))

for (i in seq_along(cols)) {
  sd_vec[i] <- sd(nyc[[cols[i]]], na.rm = TRUE)
}

names(sd_vec) <- cols
sd_vec

```


>- Use ifelse to create a new variable of high and normal A1C dichotomizing with 7 as a threshold for high levels

```{r}

A1C_levels <- ifelse(nyc$A1C >= 7, "High", "Normal")

```


>- Subset to only have females (gender == "female")

```{r}
nyc_females <- subset(nyc, GENDER == 2)

```


>- Create a subset of males and find the median total cholesterol level

```{r}
nyc_males <- subset(nyc, GENDER == 1)
median(nyc_males$CHOLESTEROLTOTAL, na.rm =TRUE)
```


>- Try to create a summary of the mean A1C using two factors, GENDER and DX_DBTS

```{r}

aggregate(A1C ~ GENDER + DX_DBTS, data = nyc, FUN = mean, na.rm = TRUE)

```

Find the class of the first three columns of data. What are they?

```{r}

sapply(nyc[, 1:3], class)

```

